Forcing for Isabelle

### 2018-12-02 LICS/forcing.tex
ispell y otras cosas
trayendo cosas de blueprint sin ningún reparo

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% End:

>>>>

\subsection{The fundamental theorems}
These are stated as two lemmas: \emph{Truth} and
\emph{Definability}. The second one ``simply'' says that the forcing
relation is definable. The previous recursive construction amounts to
that. The Truth Lemma states that the forcing relation indeed relates
satisfaction in $M[G]$ to that in $M$. 
\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
  language of set 
  theory with all free variables displayed. Then the
  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
  \begin{enumerate}
  \item $M[G],[\val(G,\tau_0),\dots,\val(G,\tau_n)]
  \models\phi(x_0,\dots,x_n)$ for 
    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
  \item $  M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n] 
  \models\forceisa(\phi(x_4,\dots,x_{n+4}))$.
  \end{enumerate}
\end{lemma}
The proof of the Truth Lemma is by recursion in elements of
$\formula$. The following auxiliary results (adapted from
\cite[IV.2.43]{kunen2011set}) are also proved by recursion in
$\formula$.
\begin{lemma}[Strenghtening]\label{lem:strengthen} 
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
  $M, [\PP,\leq,\1,p]\models\forceisa(\phi)$ and $p_1\leq p$
  implies $M, [\PP,\leq,\1,p_1]\models\forceisa(\phi)$.
\end{lemma}
\begin{lemma}[Density]\label{lem:density}
  Assume the same hypothesis of
  Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ 
  if and only if 
  $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
  dense below $p$.''
\end{lemma}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-12-02 LICS/proof-of-separation.tex
trayendo cosas de blueprint sin ningún reparo

To show that   
\[
S\defi\{x\in c : M[G]\models \phi(x,w)\} \in M[G],
\]
it is enough to provide a name $n\in M$ for this set.

@@

Let's first focus into the predicate 
\[
M,[u,\PP,\leq,\1,\sig,\pi]\models \psi
\]
defining $n$ by separation. By definition of the satisfaction
relation and permuting variables, we have it is equivalent to the fact
that there exist $\th,p\in M$ with   $u=\lb\th,p\rb$  and 
\[
M,[\PP,\leq,\1,p,\th,\sig,\pi]\models \forceisa(x_4\in
x_6\y\phi(x_4,x_5)). 
\]
% (Note that the variable $x_7$ is not used.)
This, in turn is equivalent by the Definition of Forcing to: For all $M$-generic
filters $F$ such that $p\in F$, 
\[
M,[\val(G,\th),\val(G,\sig),\val(G,\pi)]\models x_4\in
x_6\y\phi(x_4,x_5). 
\]

%% The proof of
%% \isatt{Collect{\isacharunderscore}sats{\isacharunderscore}in{\isacharunderscore}MG}
%% has three parts:
%% \begin{enumerate}
%% \item Definition of the name $n$;
%% \item Proving that 
%% \item 
%% \end{enumerate}

>>>>

To show that   
\[
S\defi\{x\in c : M[G],[x,w]\models \phi(x_0,x_1)\} \in M[G],
\]
it is enough to provide a name $n\in M$ for this set.

@@

Let's first focus into the predicate 
\begin{equation}\label{eq:1}
M,[u,\PP,\leq,\1,\sig,\pi]\models \psi
\end{equation}
defining $n$ by separation. By definition of the satisfaction
relation and %% permuting variables
absoluteness, we have that it is equivalent to the fact
that there exist $\th,p\in M$ with   $u=\lb\th,p\rb$  and 
\[
M,[\PP,\leq,\1,p,\th,\sig,\pi]\models \forceisa(x_4\in
x_6\y\phi(x_4,x_5)). 
\]
% (Note that the variable $x_7$ is not used.)
This, in turn, is equivalent by the Definition of Forcing to: \emph{For all $M$-generic
filters $F$ such that $p\in F$,} 
\begin{equation}\label{eq:2}
M[F],[\val(F,\th),\val(F,\sig),\val(F,\pi)]\models x_0\in
x_2\y\phi(x_0,x_1). 
\end{equation}
We can instantiate this statement with $G$ and obtain
\[
p\in G \impl M[G],[\val(G,\th),w,c]\models x_0\in
x_2\y\phi(x_0,x_1). 
\] 
Let $Q(\th,p)$ be the last statement. We have just seen that
(\ref{eq:1}) implies 
\[
\exists \th,p\in M.\ u=\lb\th,p\rb \y Q(\th,p).
\]
Hence $n$ is included in 
\[
m\defi \{u \in\dom(\pi)\times\PP : \exists \th,p\in M.\ u=\lb\th,p\rb
\y Q(\th,p)\}. 
\]

It can be seen that
\begin{lemma}
  $\val(G,m) = S$.
\end{lemma}
\noindent And by monotonicity of $\val$ we obtain
\begin{lemma}
  $\val(G,n)\sbq \val(G,m)$.
\end{lemma}
To complete the proof, it is therefore enough to show that
$S\sbq \val(G,n)$. For this, let $x\in S$. Hence there exists
$\lb\th,q\rb\in\pi$ such that 
$q\in G$ and $x=\val(G,\th)$.

On the other hand, since 
\[
M[G],[\val(G,\th),\val(G,\sig),\val(G,\pi)]\models
 x_0\in x_2\y\phi( x_0, x_1),
\]
by the  Truth Lemma there must exist $r\in G$ such that
\[
M,[\PP,\leq,\1,r,\th,\sig,\pi]\models
\forceisa(x_4\in x_6\y\phi( x_4, x_5)).
\]
Since $G$ is a filter, there is $p\in G$ such that $p\leq q, r$.
By Strenghtening, we have
\[
M,[\PP,\leq,\1,p,\th,\sig,\pi]\models
\forceisa(x_4\in x_6\y \phi( x_4, x_5)),
\]
which by the Truth Lemma gives us: \emph{for all $M$-generic $F$,
  $p\in F$ implies} 
\[
M[F],[\val(F,\th),\val(F,\sig),\val(F,\pi)]\models
 x_0\in  x_2 \y\phi( x_0, x_1).
\]
Note this is the same as (\ref{eq:2}). In fact, we have that
\begin{lemma}
  $\val(G,n)$ equals
  \[
  \{\val(G,\th): \th\in\dom(\pi) \y  p \in G 
  \y \text{(\ref{eq:2}) holds}\}.
  \]
\end{lemma}
This shows that $S\sbq \val(G,n)$, since $\th\in\dom(\pi)$ by construction.
